|

### 2020-06-30 paper.Rmd
Working on the introduction.

Talk about:

- linear projections and tours [@As85,@BCAH05]: we want to discover features, by looking at the data from all sides
- most projections are approximately gaussian, as dimensionality increases, observations pile in the center [@diaconis1984]
- related: crowding problem as addressed in t-SNE [@tsne]
- connection with high-dimension low-sample data? observations will be away from the center, on the vertices of a simples [@doi:10.1111/j.1467-9868.2005.00510.x], and we can find directions where data is piling in a single point [@10.2307/27639976,@10.1093/biomet/asp084]
- for modelling, we often address high-dimensional problems with regularisation methods (ridge regression, Lasso). These are optimisation techniques that shrink the parameter estimates (ridge regression gives global shrinkage, Lasso can be thought of as variable selection and will shrink some of the estimates to zero). Our approach is complementary, we still consider all variables and possible combinations, but move the focus away from the extreme values, instead focusing in towards the center, where observations tend to pile up.

We have to think about the low-d projections differently than the full distribution. As $p$ increases the full volume of the parameter space grows exponentially, and we get a low density of points, and in general the center of the distribution will be empty [@doi:10.1111/j.1467-9868.2005.00510.x].

@@

Similarly, sections of the data will typically capture only a very small fraction of the points. However, when making projections, we are effectively marginalising the distribution over all orthogonal directions, and this averaging often leads to the piling of observations in the center.

>>>>

The notion of a curse of dimensionality was originally introduced in @BellmanRichard1961 to describe the increasing difficulty of optimisation in high dimensions when evaluating the objective function on a grid. More generally, we can think of it as a problem of sampling large parameter spaces. As the number of dimensions increases, the volume of the parameter space grows exponentially, and the density of points decreases. As a result observations will be far from the center of the distribution, on the vertices of a simplex [@doi:10.1111/j.1467-9868.2005.00510.x]. This causes problems in different domains, an important example in statistics is cluster analysis: since the distance between any two points in a distribution becomes approximately constant.

A more detailed discussion of the curse of dimensionality was presented in @Donoho00, which also highlighted the opposite effect, a blessing of dimensionality. Indeed we can sometimes exploit large parameter spaces to our advantage. One example is the use of regularisation methods for variable selection. These methods penalise model complexity in the optimisation, for example ridge regression and Lasso. The penalty results in shrinking (some of) the parameter estimates towards zero. In ridge regression we obtain a global shrinkage, while Lasso can be thought of as variable selection and will shrink some of the estimates to zero.

A different view on the curse of dimensionality emerges for dimension reduction and visualisation methods. When working with low-dimensional linear projections from a high-dimensional parameter space, most projections will show approximately Gaussian distributions of the data, and as dimensionality increases, observations start to pile near the center [@diaconis1984]. In the extreme case of high-dimension low-sample size data this results in "data piling", and we can find directions where all observations are in a single point [@10.2307/27639976,@10.1093/biomet/asp084]. These issues also persist with non-linear dimension reduction techniques. This is often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.

To understand these seemingly contradictory implications, low density of points near the center in the high-dimensional space, and the piling in low-dimensional projections, we need to consider the projected volume across the low-dimensional projection. We can think of the data being distributed in a hypersphere, i.e. we think of a distribution with a maximum distance from the center. Compared to the common assumption of a distribution in a box or hypercube this has the advantage of being rotation invariant, and more accurately captures commonly observed distributions. We can picture the linearly projected volume starting from a 3D sphere. In a 2D projection, a large fraction of the volume will be projected onto a small region near the center of the resulting circle, and will be even further concentrated when projecting onto 1D. The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualisation, with increasing fractions of the volume being projected onto a small area around the center. The piling of observaitions then follows directly by assuming that they are uniformly distributed across the volume.

In this work we address this piling in low-dimensional linear projections, which is in particular a barrier for viewing high-dimensional distributions with tour methods [@As85,@BCAH05]. Tours show interpolated sequences of low-dimensional projections of the data. When exploring a dataset with a tour we look at the distribution from different sides and want to discover features that may be visible from a specific viewing angle. However, these features may easily be hidden if most of the observations pile up near the center.  XXX talk about few outliers taking up large part of the plotting space

@@
